Density-dependent selection and the limits of relative fitness

### 2017-12-14 manuscript.tex
l

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time $t$ advances in discrete iterations, each representing the time from birth to reproductive maturity. In iteration $t$, the number of adults of the $i$'th type is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.

@@

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
\end{equation}
in our extended lottery model, where the sum only includes territories with at least one propagule present.

Finally, we assume that mortality only occurs in adults (Fig.~\ref{fig:lottery}; setting aside the juvenile deaths implicit in territorial contest), and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in type abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}

@@

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type $l_i\gg 1$. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Then the numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively. Type $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving 
\begin{equation}
\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=\frac{c_i l_i}{\overline{c}L}U(t), \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.

Eq. \eqref{eq:lottery} breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density. Instead, the propagules from a low density type only contest a small fraction of territories, and do so with at least one propagule present (not a small fraction of a propagule $l_i\ll 1$). In section ``Analytical approximation of the density-dependent lottery'', we present a generalization of \eqref{eq:lottery} that accounts for this low density behavior.

There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}. In the Wright-Fisher model, type abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  type abundances in the preceding generation. Population size $N$ remains constant. This is equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and relative fitness given by $w_i=b_i c_i$ i.e. a product of fecundity and viability \citep[pp. 185]{crow_1970}. Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift. This means that our extension of the classic lottery model to arbitrary densities represents a density-dependent generalization of the Wright-Fisher model.

@@

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U(t), \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}. \nonumber \label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}. The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; total population density thus obeys 
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

@@

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant during a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical (and hence so is saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b). Selection then occurs purely along this line, uncoupled from the regulation of density. In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all. While quite different from classical growth models like the Lotka-Volterra, this is all perfectly consistent with MacArthur's general argument.

@@

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong. To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}. The saturation density in Eq.~\eqref{eq:simplebirthdeath} is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over the $\delta$-sweep. Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.1$ and upwards (Fig.~\ref{fig:strengthofselection}).

The cause of this effect is that the mortality rate $\delta_i N$ is both density-regulating and subject to density-dependent selection; thus selection on $\delta$ induces a change in $N$ but also changes in strength in response to $N$.

@@

Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:simplebirthdeath} with our density-dependent lottery. As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps. For $b$ and $d$ sweeps, $s$ is also exactly constant, because the overall regulation of population density applies equally to all types \citep{prout_1980}. To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
\begin{equation}
\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
\end{equation}
Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$. This applies uniformly to all types and therefore does not induce density-dependence in selective advantages related to $b$ or $d$. Note that our lottery model accounts for two key aspects of the interaction between selection and demography: selection is density-dependent, and density will change as a result of selection on absolute fitness traits. And yet pure $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly. [Need to flesh this out a little more and address some subtleties with discrete vs. continuous time]

@@

>>>>

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time advances in discrete iterations, each representing the time from birth to reproductive maturity. In a given iteration the number of adults of the $i$'th type will be denoted $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.

@@

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
\end{equation}
in our extended lottery model, where the sum only includes territories with at least one propagule present.

Finally, we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in type abundances is
\begin{equation}
\Delta n_i=\Delta_+ n_i-d_i n_i. \label{eq:delttot}
\end{equation}

@@

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type $l_i\gg 1$. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Then the numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively. Type $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving 
\begin{equation}
\Delta_+ n_i=\frac{c_i l_i}{\sum_j c_j l_j}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules. In section ``Analytical approximation of the density-dependent lottery'', we derive a generalization of Eq.~\eqref{eq:lottery} that accommodates arbitrary propagule densities $l_i$.

There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}. In the Wright-Fisher model, type abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  type abundances in the preceding generation. Population size $N$ remains constant. This is equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and relative fitness given by $w_i=b_i c_i$ i.e. a product of fecundity and viability \citep[pp. 185]{crow_1970}. Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift. This means that our extension of the classic lottery model to arbitrary densities represents a density-dependent generalization of the Wright-Fisher model (we also do not consider drift here).

@@

Similarly to the high-$l_i$ approximation of the classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$ as in Eq.~\eqref{eq:lottery}. The classic lottery model breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density. For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U, \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}. \nonumber \label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population (i.e. it determines invasion fitness \citep{metz_1992}). The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).

@@

Eq.~\eqref{eq:master} takes a much simpler form if all types are competitively equivalent ($c_i=c$),
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
\end{equation}
Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives $l_i/L$ of these. Total population density thus grows according to
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but the types are otherwise identical (so that each type has the same saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b). Selection then occurs purely along this line, similarly to Fig.~\ref{fig:Ksel}b. This occurs because $c$ does not directly affect $N$: it only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place (this can be seen formally in Eq.~\eqref{eq:Nmaster}). In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. While quite different from classical growth models like the Lotka-Volterra, this is all perfectly consistent with MacArthur's general argument.

@@

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong. To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$. The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.1$ and upwards (Fig.~\ref{fig:strengthofselection}).

@@

Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:simplebirthdeath} with our density-dependent lottery. Selection on $c$ is clearly density-dependent, but since selection on $c$ does not affect density (section ``$K$-selection versus relative fitness''), selection on $c$ does not introduce density-dependence in Eq.~\eqref{eq:canonical}.

Turning to selection on $b$ and $d$, recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
\begin{equation}
\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
\end{equation}
The mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent, and so $d$ sweeps also follow the canonical relative fitness model exactly.

At first glance $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$. This term declines from $\overline{b}$ at low density to zero at high density and as a result, selection on $b$ is density-dependent. In words: the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.

Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above. The reason is that $b$ does not regulate density via the $b_i$ term in front of $f(\overline{b},N)$, which is divided by $\overline{b}$ and thus actually reflects relative lottery contests. The density regulating effect of $b$ actually appears in the fraction of contested territories $1-e^{-\overline{b}N/T}$. This reflects the fact that $b$ sweeps can only increase density by causing more unoccupied territories to receive propagules. The net effect on $f(\overline{b},N)$ is precisely zero. To see this, note that in a single-type equilibrium we have $b_i/\overline{b}=1$ and so, setting Eq.~\eqref{eq:bdensitydependence} to zero, we $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases. Strictly speaking there is some deviation in $f(N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_i\rightarrow b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details). Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we can safely conclude that $b$ sweeps obey Eq.~\eqref{eq:canonical} to a very close approximation.

To summarize: (i) $c$-selection is density-dependent, but $c$ does not affect density (ii) $d$-selection is density-independent but $d$ affects density (iii) $b$ affects density and is subject to density-dependent selection. And yet pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation exactly (almost exactly for $b$).

@@

\section*{Appendix D: Density-dependence of $b$-selection}

In section ``Density-dependence and the strength of selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$. Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.

For simplicity, we introduce the notation $D=N/T$ and assume that $D$ is small. We can thus make the approximation $1-e^{-\overline{b}D}\approx \overline{b}D$ and $f(\overline{b},N)\approx \overline{b}(1-D)$. We expect this to be a conservative approximate based on the worst case scenario, because $N$ is most sensitive to an increase in $b$ in this low-density linear regime. We first calculate the value of $f(\overline{b},N)$ at the halfway point in a sweep, where the halfway point is estimated with simple linear averages for $b$ and $N$. The sweep is driven by a $b$ variant with $b_j=b_i(1+\epsilon)$, and we denote the corresponding initial and final densities by $D_i$ and $D_j$ respectively, where we have $d_i=b_i(1-D_i)=b_j(1-D_j)$. We obtain
\begin{align}
f_{\rm{half}}=f(\frac{b_i+b_j}{2},\frac{N_i+N_j}{2})&=\frac{b_i+b_j}{2}\left(1-\frac{D_i+D_j}{2}\right) \nonumber \\
&=\frac{1}{4} (b_i+b_j)(2-D_i-D_j) \nonumber \\
&=\frac{1}{4} (2d_i+b_i(1-D_j)+b_j(1-N_i)).
\end{align}
Dividing by $d_i$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
\begin{align}
\frac{f_{\rm{half}}}{d_i}&=\frac{1}{4}\left(2+\frac{b_i}{b_j}+\frac{b_j}{b_i}\right)\nonumber \\
&=\frac{1}{4}\left(2+\frac{1}{1+\epsilon}+1+\epsilon\right)\nonumber \\
&=1+\frac{1}{4}(\epsilon^2-\epsilon^3+\ldots),
\end{align}
where we have used the Taylor expansion $\frac{1}{1+\epsilon}=1-\epsilon+\epsilon^2-\epsilon^3+\ldots$.

By contrast, for a $\delta$ sweep in Eq.~\eqref{eq:simplebirthdeath}, the density-dependent term $N$ increases by a factor of $\frac{1}{1-\epsilon}=1+\epsilon+\epsilon^2+\ldots$. Thus,  the deviations in $f(N)$ are an order of magnitude smaller than those shown in Fig.~\eqref{fig:strengthofselection}, and even proportional changes of order $\epsilon=0.1$ will cause a negligible deviation from the canonical selection equation.
